\begin{slide}{Motivation of Soot Attributes in Eclipse}
\begin{itemize}
\item The Soot - Eclipse plugin provides a mechanism for viewing
attribute information in visual ways within Eclipse.
\item This can aid in software visualization, program understanding and analysis debugging.
\end{itemize}
\end{slide}

\begin{slide}{Visual Representations}
\begin{itemize}
\item There are currently three different visual representations of attribute informatation within the plugin.
\begin{itemize}
\item Text is displayed in tooltips over a statement.
\item Chunks of code can be highlighted with different colors.
\item Links are displayed in a popup list and reposition the cursor when selected.
\end{itemize}
\end{itemize}
\end{slide}

\begin{slide}{String Tags}
\begin{itemize}
\item {\red StringTag}s attach a string of information to a {\red Host}.
{\scriptsize
\begin{verbatim}
s.addTag(new StringTag(val+": NonNull"));
\end{verbatim}
}
\item The Soot - Eclipse plugin displays the string as a tooltip when the mouse hovers over a line of text in the Java editor and Jimple editor.
\end{itemize}
\sablefootnote{soot.tagkit.StringTag}
\end{slide}

\begin{slide}{Color Tags}
\begin{itemize}
\item {\red ColorTag}s attach a color to a {\red Host}.
{\scriptsize
\begin{verbatim}
v.addTag(new ColorTag(ColorTag.GREEN));
v.addTag(new ColorTag(255, 0, 0));
\end{verbatim}
}
\item The Soot - Eclipse plugin highlights the background color of the text in the editor at the appropriate positions with the given color in the Jimple editor.
\end{itemize}
\sablefootnote{soot.tagkit.ColorTag}
\end{slide}

\begin{slide}{Link Tags}
\begin{itemize}
\item {\red LinkTag}s attach a string of information, and a link to another part of code to a {\red Host}.
{\scriptsize
\begin{verbatim}
String text = "Target:"+m.toString();
Host h = m;
String cName = m.getDeclaringClass().getName();
s.addTag(new LinkTag(text, h, cName));
\end{verbatim}
}
\item The Soot - Eclipse plugin displays link which jumps to a another part of the code when clicked in the Jimple Editor.
\end{itemize}
\sablefootnote{soot.tagkit.LinkTag}
\end{slide}

